Experience Report: Haskell in Computational Biology

### 2012-03-09 mrfy_experience_report.tex
trying to restore parallel struture in the intro

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
For some software, considerations of performance are paramount; this
software is usually written in C~or~C++. 
For other software, considerations of convenience, readability, and
productivity are more important;
this software is usually written in a ``scripting'' language like
Perl, Python, Ruby, or~R.\remark{Don't forget audience.}
In~this Experience Report, we show that computational biologists can
successfully replace both these families of languages with Haskell:
\begin{itemize}
\item
Our experimental Haskell programs are as easy to understand and change
 as the programs our group writes using scripting languages.
\item
The performance of our Haskell code is comparable to the performance
of~C++.
\item
Although we have had to write a new implementation of a standard
algorithm, the cost was low, and the connection between code and
algorithm is unusually strong (\secref{viterbi}).
\item
We have experienced barriers to entry, of which the most salient has been
the difficulty of understanding and tuning  the performance of Haskell
programs. 
We found the other barriers 
relatively easy to overcome.
\end{itemize}
Our experience is based on MRFy (pronounced ``Murphy''), which
implements a novel family of algorithms for finding remote homologs to 
newly discovered proteins. 
\remark{``Homology'' means ``share a common ancestor.''  I hope we can
avoid this story. ---NR}

@@

Proteins are cellular machinery. They interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signalling, and more.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
\remark{Lost: there are only 20-odd amino acids}
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a genetic sequence (the \textit{query sequence}), biologists wish to know the cellular
function of the protein that gene codes for.
\remark{I thought the query sequence was an amino-acid sequence.}
The best known method of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
Proteins that share structure and function are \textit{homologous}, and thus
the process of identifying similar proteins to a query sequence is called 
\textit{homology detection}.

Computational biologists detect homologies by building 
algorithms which, given an amino-acid query sequence,
find known proteins of similar structure.
When the similar proteins are formed from amino-acid sequences that
are nearly identical to the sequence under test, the homologies can be
detected by  
simple string-matching algorithms.\remark{Can we just drop the
  sentence about string matching?}
When the similar proteins are formed from amino-acid sequences that
are not too far in edit distance from the sequence under test, the homologies can be
detected by
a family of algorithms called 
\textit{hidden Markov models}~\cite{Eddy:1998ut}.
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research contribution---the MRFy algorithm---can detect homologies
in amino-acid sequences that are only distantly related.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.

\emph{[We need an argument here saying why an understanding of the
    algorithms is relevant to the claims made in the experience
    report, then say how we present the algorithms, then say what
    experiential conclusions this presentation supports.]}

We present an explanation of the algorithms relevant to MRFy, because there 
exists a natural fit between the mathematics behind the algorithms and the 
Haskell implementation of those algorithms.
If the reader understands the 
algorithms, he or she is more likely to appreciate the benefits of implementing 
them in Haskell.
We will first present the more traditional algorithm used in 
hidden Markov models, followed by our variation on it and the stochastic search 
approach.
We conclude that implementations of mathematically-intensive 
algorithms such as these lend themselves to the Haskell language, particularly 
when the programmer needs a flexible implementation that simplifies 
experimenting with variants of the algorithm.

@@

Homology-detection software most often used in one of two ways:
to test a hypothesis about 
the function of a single, newly discovered protein, or 
to compare every protein in a genome against a library of known protein 
structures.
The software is ``trained'' \remark{do we want to talk ``Training''?}
on a group of proteins that share function and structure.
These proteins are identified by a biologist, who puts
their amino-acid sequences into a correspondence, also called an
\emph{alignment}. This alignment signifies structural and evolutionary
correspondence among individual amino acids in a set of homologous proteins.
An~alignment may be represented as a matrix
(Figure~\ref{alignment}) 
in which each row corresponds to the amino-acid sequence of a protein,
and each column groups amino acids that play similar roles in
different proteins.
A protein may have a gap in a column of the alignment, which indicates
that during the course of evolution, either that protein lost that amino
acid, or other proteins gained an amino acid.
If~a column contains just one or two gaps, the most likely explanation
is that in one or two proteins, mutations in the underlying DNA
deleted an amino acid from that position (\figref{alignment}).
If~a column contains \emph{mostly} gaps, 
the most likely explanation
is that in one or two proteins, mutations in the underlying DNA
\emph{inserted} an amino acid in that position (\figref{alignment}).

The protein alignment is used to train a \emph{hidden Markov model}.
\remark{I've never seen this hypenated, in ANY literature.}
A~hidden Markov model is a probabilistic finite-state machine which 
can assign a probability to any query sequence of amino acids.
Proteins whose sequences have higher probabilities are more likely to
be homologous to the proteins in the alignment.
Mathematically, we represent a query sequence as
$\left\{x_{1}...x_{n}\right\}$.
Ideally, if we were writing something like the BioHaskell library, we would
have begun with an algebraic datatype for amino acids deriving
\texttt{Ix}, something like
\begin{verbatim}data Amino = Ala | Cys | Asp | Glu | Gly ...\end{verbatim}
which would still allow us to index into an array or vector.
However, since we wanted to maintain compatibility with the file format
shared with previous software, we broke with idiomatic Haskell and represented
our amino acids as 
\begin{verbatim}
type AA = Int
\end{verbatim}
and thus our query sequence as
\begin{verbatim}
type querySeq = Vector AA
\end{verbatim}

The states of the hidden Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
If a given column~$j$ of the alignment contains amino acids in a majority
of the protein sequences, it is considered a \textit{consensus column}.
\item
For a given column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}.
The match state determines the probability that the query sequence
has a particular amino acid in column~$j$.
The match state assigns high probability to the amino acids that
appear in column~$j$, and a low but nonzero probability to other amino
acids.
The match state is more probable if it corresponds to a consensus column.
\item
For every column of the alignment, the hidden Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$j$, and low probability if there are few gaps in column~$j$.
Thus, the deletion state is more likely in non-consensus columns.
\item
For every column of the alignment, the hidden Markov model has an
\emph{insertion state}.
The insertion state determines the probability that a query sequence
has an ``extra'' amino acid that does not correspond to any
consensus column in the alignment.
Thus, the insertion state is more likely in non-consensus columns.
The insertion state assigns high probability to amino acids that are
frequently observed among all proteins, and low probability to amino acids
that are rarely observed.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
The probability associated with a specific state is conditional upon the previous
state in the model, so it is a \textit{transition probability}.
Mathematically, transition probabilities are represented by $a_{s_{j-1}s_{j}}$
for a state $s_{j}$.
In Haskell, we use 
\begin{smallverbatim}
type logProb = Double
data TProb  = TProb { logProbability::LogProb
                    , fromState::HMMState
                    , toState::HMMState
                    }
\end{smallverbatim}
to represent the seven allowed state transitions.

>>>>

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
For some software, considerations of performance are paramount; this
software is usually written in C~or~C++. 
For other software, considerations of convenience, readability, and
productivity are more important;
this software is usually written in a ``scripting'' language like
Perl, Python, Ruby, or~R.\remark{Don't forget audience.}
In~this Experience Report, we show that computational biologists can
successfully replace both these families of languages with Haskell:
\begin{itemize}
\item
Our experimental Haskell programs are as easy to understand and change
 as the programs our group writes using scripting languages.
\item
The performance of our Haskell code is comparable to the performance
of~C++.
\item
Although we have had to write a new implementation of a standard
algorithm, the cost was low, and the connection between code and
algorithm is unusually strong (\secref{viterbi}).
\item
We have experienced barriers to entry, of which the most salient has been
the difficulty of understanding and tuning  the performance of Haskell
programs. 
We found the other barriers 
relatively easy to overcome.
\end{itemize}
Our experience is based on MRFy (pronounced ``Murphy''), which
implements a novel family of algorithms for finding remote homologs to 
newly discovered proteins.

@@

Proteins are cellular machinery. They interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signalling, and more.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
\remark{Lost: there are only 20-odd amino acids}
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a genetic sequence, biologists wish to know the cellular
function of the protein that gene codes for.
The best known method of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
Proteins that share structure and function are expected to be
descended from a common ancestor---in biological terms, \emph{homologous}---and
thus
the problem of identifying proteins similar to a query sequence is called 
\textit{homology detection}.

Computational biologists detect homologies by building 
algorithms which, given a \emph{query sequence} of amino acids,
find known proteins of similar structure.
When the similar proteins are formed from amino-acid sequences that
are not too different from the query sequence, the homology can be
detected by
a family of algorithms called 
\textit{hidden Markov models}~\cite{Eddy:1998ut}.
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research contribution---the MRFy algorithm---can detect homologies
in amino-acid sequences that are only distantly related.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.
%
To~enable you to understand our experience with Haskell,
we~sketch not only the MRFy algorithm (\secref{mrfy})
but also one of the older hidden-Markov algorithms (\secref{viterbi}),
which is incorporated into~MRFy.

\remark{LOST: We conclude that implementations of mathematically-intensive 
algorithms such as these lend themselves to the Haskell language, particularly 
when the programmer needs a flexible implementation that simplifies 
experimenting with variants of the algorithm.}

@@

Homology-detection software most often used in one of two ways:
to test a hypothesis about 
the function of a single, newly discovered protein, or 
to compare every protein in a genome against a library of known protein 
structures.
Either way, 
the software is \emph{trained}
on a group of proteins that share function and structure.
These proteins are identified by a biologist, who puts
their amino-acid sequences into a correspondence, also called an
\emph{alignment}. 
This alignment represents structural
correspondence among individual amino acids in a set of homologous proteins.
An~alignment may be represented as a matrix
(Figure~\ref{alignment}) 
in which each row corresponds to the amino-acid sequence of a protein,
and each column groups amino acids that play similar roles in
different proteins.

An alignment may contain \emph{gaps} (notated as dashes in
\figref{alignment}) for some proteins.
A~gap in row~2, column~$j$ indicates that as proteins evolved, either 
protein~2 lost its amino acid in position~$j$, or 
other proteins gained an amino acid in position~$j$.
If~column~$j$ contains few gaps, 
it~is considered a \emph{consensus column},
and the few proteins with gaps probably lost amino acids via
\emph{deletions}. 
If~column~$j$ contains \emph{mostly} gaps, 
it~is considered a \emph{non-consensus column},
and the few proteins without gaps probably gained amino acids via
\emph{insertions}.

Once a protein alignment is constructed, it~is used to train a
\emph{hidden Markov model}. 
A~hidden Markov model is a probabilistic finite-state machine which 
can assign a probability to any query sequence.
A~protein whose query sequence has a higher probability is more likely to %
be homologous to the proteins in the alignment.
We~write a query sequence as $x_1, \ldots, x_{\scriptscriptstyle N}$,
where each $x_i$~is 
an amino acid.

The states of the hidden Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}.
The match state determines the probability that the query sequence
has a particular amino acid in column~$j$.
The match state assigns high probability to the amino acids that
appear in column~$j$, and a low but nonzero probability to other amino
acids.
A~match state is more probable if it corresponds to a consensus column.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
A~deletion state is more probable if it corresponds to a
non-consensus column.
% has high probability if there are lots of gaps in
%column~$j$, and low probability if there are few gaps in column~$j$.
%Thus, the deletion state is more likely in non-consensus columns.
\item
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}.
The insertion state determines the probability that a query sequence
has an ``extra'' amino acid.
Like a match state, the insertion state assigns high probability to
amino acids appearing in column~$j$.
An~insertion state is more probable if it corresponds to a
non-consensus column.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
The probability associated with a specific state is conditional upon the previous
state in the model, so it is a \textit{transition probability}.
Mathematically, transition probabilities are represented by $a_{s_{j-1}s_{j}}$
for a state $s_{j}$.
In Haskell, we use 
\begin{smallverbatim}
type logProb = Double
data TProb  = TProb { logProbability::LogProb
                    , fromState::HMMState
                    , toState::HMMState
                    }
\end{smallverbatim}
to represent the seven allowed state transitions.

We~write a query sequence as
$x_{1}, \ldots, x_{n}$, where each~$x_i$ is an amino acid.
In~idiomatic Haskell, we might represent each individual amino acid
using a value of algebraic data type, thus:
\begin{smallverbatim}
data Amino = Ala | Cys | Asp | Glu | Gly | ...
\end{smallverbatim}
But our primary goal was to get something working quickly with a
legacy file format that represents each amino acid as a small integer.
Our types are therefore
\begin{smallverbatim}
type AA       = Int -- amino acid using HMMER coding
type QuerySeq = Vector AA
\end{smallverbatim}
